Perturbing SU($n$)-symmetric interactions

### 2019-12-06 collective_sun/sun_perturbation.tex
added appendix on a minor improvement to the second order two-body theory

\newcommand{\FS}{\text{FS}}
\newcommand{\EQFS}{=_{\text{FS}}}
\newcommand{\col}{\underline}

\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\cov}{cov}
\let\var\relax
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\diag}{diag}
\let\flat\relax
\DeclareMathOperator{\flat}{flat}
\DeclareMathOperator{\counts}{counts}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Single-body perturbations}

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_p v_p \P_0 X_p \P_0
  = \sum_p v_p \P_0 X_0 \P_0 \EQFS \bar v \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v \equiv \f1N \sum_p v_p,
  &&
  \col{X} \equiv \sum_p X_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q h_q v_q X_q} \ket\psi,
  \label{eq:H_V1_psi}
\end{align}
where $h_q\equiv\sum_p h_{pq}$ is the net interaction energy
associated with spin $p$.  We thus find that the vector $\V_1\ket\psi$
is an eigenvector of the interaction Hamiltonian $H_0$ if the
coefficients $\v v$ satisfy
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta\p{\v v} \v v,
  \label{eq:cond_1}
\end{align}
where $\v h,\m h$ are respectively a vector and matrix of all
$h_p,h_{pq}$; $\diag\v h$ is a matrix with $\v h$ on the diagonal and
zeros everywhere else; and $\Delta\p{\v v}$ is some constant that may
depend on $\v v$.  If we {\it construct} a perturbation
$\V_1\p{\v v_\Delta}$ with coefficients $\v v_\Delta$ that satisfy
\eqref{eq:cond_1} with eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v_\Delta}\ket\psi$ depends only on the coefficients
$\v v_\Delta$, and is entirely independent of the state $\ket\psi$ or
choice of single-spin trace-zero operator $X$ used to build $\V_1$ in
\eqref{eq:perturbations}.  Finding operators $\V_1\p{\v v_\Delta}$
that generate eigenvectors of the interaction Hamiltonian $H_0$ when
they act on fully symmetric states $\ket\psi$ thus reduces to finding
eigenvectors $\v v_\Delta$ of the interaction matrix $\m h-\diag\v h$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two-body perturbations}

@@

We can recast the condition in \eqref{eq:cond_2} as an eigenvalue
problem by writing the coefficient matrix $\m w$ as a vector
$\v{\m w}$ in $\RR^N\times\RR^N$,
\begin{align}
  \m w = \sum_{p,q} w_{pq} \op{p}{q}
  \to \v{\m w} \equiv \flat\m w \equiv \sum_{p,q} w_{pq} \ket{pq},
\end{align}
where $\flat\m m$ is a flattening of the matrix $\m m$ into a vector
$\vec{\m m}$, and defining the matrix
\begin{align}
  \check{\m h}
  \equiv 2\diag\p{\flat\bar{\m h}}
  + \m h\otimes\1_N + \1_N\otimes\m h
  = 2 \sum_{p,q} \bar h_{pq} \op{pq}
  + \sum_{p,q,\ell} h_{pq} \p{\op{p\ell}{q\ell} + \op{\ell p}{\ell q}},
  \label{eq:h_super_mat}
\end{align}
where $\1_D$ is a $D\times D$ identity matrix.  The condition in
\eqref{eq:cond_2} is then equivalently the eigenvalue equation
\begin{align}
  \p{\check{\m d} \c \check{\m h} \c \check{\m d}} \c \v{\m w}
  = \Delta\p{\m w} \v{\m w},
  &&
  \check{\m d} \equiv \1_{N^2} - \sum_\ell \op{\ell\ell},
  \label{eq:cond_2_eig}
\end{align}
where $\check{\m d}$ annihilates the diagonal part of a (flattened)
matrix it acts on, while keeping the off-diagonal components the same.
We conjugate $\check{\m h}$ by $\check{\m d}$ to enforce that its
eigenvectors $\v{\m w}$ correspond to matrices $\m w$ with zeros on
the diagonal, i.e.~$w_{pp}=0$.

Both numerical experiments and physical arguments suggest that the
$N^2\times N^2$ matrix $\check{\m d} \c \check{\m h} \c \check{\m d}$
is highly degenerate, which means that we might be able to simplify
\eqref{eq:cond_2_eig} even further.  We defer this simplification to
future work, and for now bear the (large) computational cost of
diagonalizing a $N^2\times N^2$ matrix.

@@

\appendix

@@

>>>>

\newcommand{\FS}{\text{FS}}
\newcommand{\EQFS}{=_{\text{FS}}}
\newcommand{\col}{\overline}

\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\cov}{cov}
\let\var\oldvar
\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\diag}{diag}
\let\flat\oldflat
\DeclareMathOperator{\flat}{flat}
\DeclareMathOperator{\counts}{counts}
\newcommand{\flatvec}{\underline\flat\,}
\newcommand{\ul}{\underline}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Single-body perturbations}
\label{sec:single_body}

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_p v_p \P_0 X_p \P_0
  = \sum_p v_p \P_0 X_0 \P_0 \EQFS \bar v \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQFS$ denotes equality under a restriction to the fully
symmetric manifold $\M_0$, and
\begin{align}
  \bar v \equiv \f1N \sum_p v_p,
  &&
  \col{X} \equiv \sum_p X_p.
\end{align}
In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \p{\sum_{p,q} h_{pq} v_p X_q - \sum_q h_q v_q X_q} \ket\psi,
  \label{eq:H_V1_psi}
\end{align}
where $h_q\equiv\sum_p h_{pq}$ is the net interaction energy
associated with spin $p$.  We thus find that the vector $\V_1\ket\psi$
is an eigenvector of the interaction Hamiltonian $H_0$ if the
coefficients $\v v$ satisfy
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta\p{\v v} \v v,
  \label{eq:cond_1}
\end{align}
where $\v h,\m h$ are respectively a vector and matrix of all
$h_p,h_{pq}$; $\diag\v h$ is a matrix with $\v h$ on the diagonal and
zeros everywhere else; and $\Delta\p{\v v}$ is some constant that may
depend on $\v v$.  Solving the eigenvalue problem in \eqref{eq:cond_1}
requires diagonalizing an $N\times N$ matrix.  When the interaction
Hamiltonian $H_0$ is translationally invariant, this eigenvalue
problem can be solved analytically, which we discuss further in
Appendix \ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_1\p{\v v_\Delta}$ with
coefficients $\v v_\Delta$ that satisfy \eqref{eq:cond_1} with
eigenvalue $\Delta$, then
\begin{align}
  H_0 \V_1\p{\v v_\Delta} \ket\psi
  = \p{E_0 + \Delta} \V_1\p{\v v_\Delta} \ket\psi.
\end{align}
Interestingly, the energy of the (unnormalized) state
$\V_1\p{\v v_\Delta}\ket\psi$ depends only on the coefficients
$\v v_\Delta$, and is entirely independent of the state $\ket\psi$ or
choice of single-spin trace-zero operator $X$ used to build $\V_1$ in
\eqref{eq:perturbations}.  Finding operators $\V_1\p{\v v_\Delta}$
that generate eigenvectors of the interaction Hamiltonian $H_0$ when
they act on fully symmetric states $\ket\psi$ thus reduces to finding
eigenvectors $\v v_\Delta$ of the interaction matrix $\m h-\diag\v h$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two-body perturbations}
\label{sec:two_body}

@@

We can recast the condition in \eqref{eq:cond_2} as an eigenvalue
problem by writing the coefficient matrix $\m w$ as a vector
$\v{\m w}$ in $\RR^N\times\RR^N$,
\begin{align}
  \m w = \sum_{p,q} w_{pq} \op{p}{q}
  \to \v{\m w} \equiv \flat\m w \equiv \sum_{p,q} w_{pq} \ket{pq},
\end{align}
where $\flat\m m$ is a flattening of the $N\times N$ matrix $\m m$
into an $N^2$-component vector $\v{\m m}$, and defining the
$N^2\times N^2$ matrix
\begin{align}
  \check{\m h}
  \equiv 2\diag\p{\flat\bar{\m h}}
  + \m h\otimes\1_N + \1_N\otimes\m h
  = 2 \sum_{p,q} \bar h_{pq} \op{pq}
  + \sum_{p,q,\ell} h_{pq} \p{\op{p\ell}{q\ell} + \op{\ell p}{\ell q}},
  \label{eq:h_super_mat}
\end{align}
where $\1_D$ is a $D\times D$ identity matrix.  The condition in
\eqref{eq:cond_2} is then equivalently the eigenvalue equation
\begin{align}
  \p{\check{\m d} \c \check{\m h} \c \check{\m d}} \c \v{\m w}
  = \Delta\p{\m w} \v{\m w},
  &&
  \check{\m d} \equiv \1_{N^2} - \sum_\ell \op{\ell\ell},
  \label{eq:cond_2_eig}
\end{align}
where $\check{\m d}$ annihilates the diagonal part of a (flattened)
matrix it acts on, while keeping the off-diagonal components the same.
We conjugate $\check{\m h}$ by $\check{\m d}$ to enforce that its
eigenvectors $\v{\m w}$ correspond to matrices $\m w$ with zeros on
the diagonal, i.e.~$w_{pp}=0$.  Solving the eigenvalue equation in
\eqref{eq:cond_2_eig} nominally requires diagonalizing an
$N^2\times N^2$ matrix; we provide a reduction of this problem to that
of diagonalizing an ${N\choose2}\times{N\choose2}$
($\approx N^2/2\times N^2/2$ when $N\gg1$) matrix in Appendix
\ref{sec:reduction}.  We also discuss how to reduce this problem even
further in the case of a translationally invariant interaction
Hamiltonian $H_0$ in Appendix \ref{sec:trans_inv}.

@@

\newpage
\appendix

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reducing the two-body eigenvalue problem}
\label{sec:reduction}

At second order, the two-body perturbation theory developed in Section
\ref{sec:two_body} nominally requires diagonalizing an $N^2\times N^2$
matrix.  Here we discuss the reduction of this problem to that of
diagonalizing an ${N \choose 2} \times {N \choose 2}$
($\approx N^2/2\times N^2/2$ when $N\gg1$) matrix.  The basic idea is
to remove redundancies and trivial values from the coefficient matrix
$\m w$ that is used to construct a two-body perturbation
$\V_2\p{\m w}$.  To this end, we flatten $\m w$ into a vector
$\ul{\v{\m w}}$ in $\RR^{N\choose2}$,
\begin{align}
  \m w = \sum_{p,q} w_{pq} \op{p}{q}
  \to \ul{\v{\m w}} \equiv \flatvec\m w
  \equiv \sum_{p<q} w_{pq} \ket{\set{p,q}},
\end{align}
where we label basis vectors $\ket{\set{p,q}}$ by a choice of two
distinct integers from $\set{1,2,\cdots,N}$.  We then define the
matrix
\begin{align}
  \ul{\check{\m h}} = 2\diag\p{\flatvec\bar{\m h}}
  + \sum_{p<q}
  \sp{\sum_k \p{h_{pk} \ket{\set{q,k}} + h_{qk} \ket{\set{p,k}}}}
  \bra{\set{p,q}}.
\end{align}
The condition in \eqref{eq:cond_2} is then equivalently the eigenvalue
problem
\begin{align}
  \ul{\check{\m h}} \c \ul{\v{\m w}}
  = \Delta\p{\m w} \ul{\v{\m w}}.
\end{align}
Unlike the eigenvalue problem in \eqref{eq:cond_2_eig}, there is now
no need to conjugate $\ul{\check{\m h}}$ by an operator that
annihilates the diagonal components of $\m w$, because the
${N\choose2}$-component vector $\ul{\v{\m w}}$ already contains only
the non-trivial parts of $\m w$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Making use of translational invariance}
\label{sec:trans_inv}

[Discuss simplifications that occur to the single- and two-body theory
if $H_0$ is translationally invariant.]
